---
title: "Homework-1-R-Exercises"
author: "VISHAL BHASHYAAM"
format: 
  pdf:
    toc: true
    colorlinks: true
    number-sections: true
    
editor: visual
---

## Getting to know the Data with R

### Goal:

Practice basic R commands/methods for descriptive data analysis.

## Installing required packages

```{r}
# run install.packages if package not downloaded 
if(!require("pacman"))
  install.packages("pacman")
```

```{r}
library(pacman)

p_load(dlookr,
  DMwR2,
  GGally,
  Hmisc,
  palmerpenguins,
  tidyverse
)

```

Loading data

The `|>` is the Base R pipe as opposed to the `magrittr` pipe `%>%`. The `|>` pipe can be utilized for most functions in R, while the `%>%` pipe is more restricted towards the `tidyverse`

```{r}
data(algae, package ="DMwR2")
algae |> glimpse()
```

# **Central tendency: mean, median, mode**

## Mean

```{r}
algae$a1 |>
  mean()
```

## Median

```{r}
algae$a1 |>
  median()
```

## Mode

Base R doesn't have a function for mode,

Creating a R function for mode, (works for unimodal, bimodal, multimodal data)

```{r}
Mode <- function(x, na.rm=FALSE){
  if (na.rm) x<-x[!is.na(x)]
  ux <- unique(x)
  return(ux[which.max(tabulate(match(x,ux)))])
}

algae$mnO2 |> Mode()
```

## `DMwR` `centralValue()` function:

returns the median for numerical variable, or the mode for nominal variables.

```{r}
# Numerical variable
algae$a1 |> centralValue()
```

```{r}
# Nominal variable
algae$speed |> centralValue()
```

## Statistics of spread (variation)

## Variance

```{r}
algae$a3 |> var()
```

## Standard deviation

```{r}
algae$a3 |> sd()
```

## Range

Note that this gives you both maximum and minimum values.

```{r}
algae$a4 |> range()
```

## Maximum value

```{r}
algae$a1 |> max ()
```

## Minimum value

```{r}
algae$a1 |> min()
```

## Interquartile range

3rd quartile (75%) - 1st quartile (25%)

```{r}
algae$a1 |> IQR()

```

## Quantiles

```{r}
algae$a1 |> quantile()
```

Specifying particular quantiles:

```{r}
algae$a1 |> quantile(probs = c(0.2,0.8))
```

## Missing values

```{r}
library(purrr)
#compute the total number of NA values in the given dataset

na_value <- algae %>%
  purrr::map_dbl(~sum(is.na(.))) %>%
  sum()

cat("The dataset contains ", na_value, "NA values. \n" )
```

```{r}
# Compute the number of incomplete rows in the dataset 


  incomplete_rows <- algae %>% 
  summarise_all(~!complete.cases(.)) %>%
  nrow()
```

```{r}
cat("The dataset contains ", incomplete_rows, "(out of ", nrow(algae),") incomplete rows. \n")
```

# Summaries of a dataset

## Baser R's `summary()`

```{r}
algae |> summary()
```

## `Hmisc`'s `describe()`

```{r}
data("penguins")
penguins |> Hmisc::describe()
```

GMD is the mean absolute difference between any pairs of observations. A robust dispersion measure, especially for non- normally distributed data.

## `dlookr`'s `describe()`

```{r}
penguins |> dlookr::describe()
```

## Summaries on a subset of data

`dplyr` 's `summarise()` and `summarise_all()` or use them with `select()` and `group_by()` to create summaries on subset of data. And,

`summarise()`= `summarize()`

```{r}
algae |> 
  summarize(avgNO3 = mean(NO3,na.rm=TRUE),
            medA1 = median(a1))

```

summarize_all() can be used to apply any function that produces a scalar value to any column of a data

```{r}
algae |> 
  select(mxPH:Cl) |> 
  summarize_all(list(mean,median), na.rm=TRUE)
```

```{r}
algae |> 
  select(a1:a7) |> 
  summarize_all(funs(var))
```

```{r}
algae |> 
  select (a1:a7) |> 
  summarise_all(c("min","max"))
```

## Use `summarize()` and `group_by()`

```{r}
algae |> 
  group_by(season, size) |> 
  summarize(n0bs = n(), mA7=median(a7))
```

```{r}
penguins |> 
  group_by(species) |> 
  summarize(var = var(bill_length_mm, na.rm = TRUE))
```

## Aggregating data

Useful for summary function that don't return scalar values

```{r}
penguins |>
  group_by(species) |>
  reframe(var = quantile(bill_length_mm, na.rm = TRUE))
```

`reframe()` expectsd a scalar result returned by the function, but quantile returns a vector.

Aggregating data with summarize was depreciated in `dplyr 1.1.0` , `reframe()` should be used instead.

```{r}
penguins |> 
  group_by(species) |> 
  dlookr::describe(bill_length_mm)

```

## Grouping functions

```{r}

```

```{r}

```

```{r}

```
